Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-04 WSSSPE1_report/WSSSPE1_summary_report.tex
changing bib to vancouver style
updates to structure overall, a few other changes scattered

\begin{abstract}
This document discusses the First Workshop on Sustainable
Software for Science: Practice and Experiences (WSSSPE1).  The content
is based on the workshop itself, as recorded in a set of
collaborative notes taken during the workshop, including discussion
about the keynote presentations, three themed panels that focus on
papers accepted by the workshop, other discussion, cross-cutting
issues, use cases, and conclusions.  It joins a nascent literature
seeking to understand what drives software work in science and how the
reward systems of science thereby shape the type of software work
undertaken, including the extent to which developers are motivated to
build software for the long-term, for the use of others, and whether
to work collaboratively or separately. Unique perspectives were
captured about issues such as documentation of APIs, software
deployment, and development practices; software licenses; career
paths for scientific software developers; identifiers for software
through a mechanism such as a Digital Object Identifier or
publication of a ``software paper''; and evidence of software
contribution and impact, recorded in online systems such as source
code repositories like GitHub. For truly sustainable software there
should be no endpoint, as the software products continue to be used
and useful beyond the single institution, grant, and developer or
development team.

@@

\note{papers and presentations and URLs should be references - see
  figshare-web in .bib file for URL style example}

@@

\pagebreak

\section*{Executive Summary}

\todo{1 page summary goes here}

\pagebreak

@@

Because progress in scientific research is dependent on the quality of and
accessibility to software at all levels, it is now critical to address many
challenges related to the development, deployment, and maintenance of reusable
software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists,
researchers, and students are able to learn and adopt software-related skills
and methodologies. Established researchers are already acquiring some of these
skills, and in particular a specialized class of software developers is
emerging in academic environments who are an integral and embedded part of
successful research teams. The WSSSPE workshop was intended to provide a forum
for discussion of the challenges, including both positions and experiences. The
short papers and debates have been archived to provide a basis for continued
discussion, and the workshop has fed into the collaborative writing of this
document. An estimate of 90 to 150 participants were present
throughout the day. Additional papers based on extended versions of workshop
submissions are expected. The level of interest in the workshop has led the
organizers, working with some of the submitters and attendees, to plan two
additional workshops (at the 2014 SciPy and SC14 conferences) and to turn the
workshop website into a community website that can be used as a focus for
further discussion and progress. Additionally, a minisymposium that aims to
further explore some of the key issues raised in WSSSPE is co-organized by a
WSSSPE1 participant at the 2014 Society for Industrial and Applied Mathematics
(SIAM) Annual Meeting on ``Reliable
Computational Science.''

Before the WSSSPE1 workshop took place, the organizers self-published
a paper~\cite{WSSSPE1-pre-report} to document the process of
organizing and advertising the workshop, collecting and reviewing the
papers, and putting together the agenda. Section~\ref{sec:process} is
based on that publication. The remainder of this paper is based on the
workshop itself, as documented in a set of collaborative notes taken
during the workshop~\cite{WSSSPE1-google-notes}, including discussion
about the keynote presentations (\S\ref{sec:keynotes}), the three
panels (\S\ref{sec:devel}-\ref{sec:community}), 
cross-cutting issues (\S\ref{sec:cross-cutting}),
use cases (\S\ref{sec:use-cases}), and conclusions
(\S\ref{sec:conclusions}).

The goal of this paper is to summarize the contents of the workshop papers
and discussion, and identify issues that future WSSSPE events will address.

@@

The organizers decided to list 54 of the papers as significant
contributions to the workshop, a very high acceptance rate, but one
that is reasonable, given the goal of broad participation and the fact
that the reports were already self-published. The papers were
grouped into three main categories, namely \emph{Developing and
Supporting Software}, \emph{Policy}, and \emph{Communities}. Each
subject area was associated with a panel and discussion at the
workshop.

@@

\todo{section should include what was presented and what was discussed}

@@

Communities are extremely important in software projects, and both their
building and continued engagement need attention during the project life cycle.
Several of the submitted papers discussed how communities have been built around
projects, and what is needed to enable a project to grow. Among these are public
source code hosting, mailing lists, documentation, wikis, bug trackers, software
downloads, continuous integration, software quality dashboards, and of course,
a general web presence to tie all of these things together. Questions were posed
around users not answering the questions of other users. Several participants
offered counterexamples, such as mailing lists where developers do not participate
as much due to users being more active, or whether the ``core
team'' can end up setting unrealistic expectations by doing too much. Team Geek\todo{add cite}
and Turk's paper on scaling code in the human dimension\todo{add cite}  discuss how development
lists that are welcoming to people actually have many more people contributing.

@@

\subsection*{Papers}

@@

\subsection*{Modeling Sustainability}

@@

\todo{seems like definitions will be mentioned elsewhere in the paper?}
Each of the framework papers included a definition of sustainability, with many overlaps.  In essence the papers related the question to change over time, an idea well captured in the definition used by the UK's Scientific Software Sustainability Institute: ``software you use today will be available---and continue to be improved and supported---in the future,'' as quoted in~\cite{Venters_WSSSPE}. This idea was also expressed as software that continues to serve its users~\cite{Pierce_WSSSPE}.

@@

\subsection*{Credit, Citation, Impact}

@@

\subsection*{Implementing Policy}

@@

\todo{section should include what was presented and what was discussed}

There were a number of papers categorized under the Communities banner and so
two presenters each summarized half of the submissions in this area.

\subsection{Communities Part 1}

This collection of papers was summarized by Karen Cranston.

@@

\subsection{Communities Part 2}

Nancy Wilkins-Diehr summarized 6 papers in this part.

@@

\subsection*{Papers}

\subsubsection*{Communities Part 1 Papers}

\subsubsection*{Communities Part 2 Papers}

\subsection{What are communities?}

@@

\subsection{Challenges of community}
include metrics for community success (e.g., more external
contributors than internal).
\subsection{Admirable scientific software communities}
\begin{itemize}
\item example communities
\item their (quick) origin stories.
\end{itemize}
\subsection{Resources for learning about software communities}
\begin{itemize}
\item academic fields studying communities (and software communities).
\item courses on online communities
\item need for software carpentry module on organizing communities?
\item \cite{howison_scientific_2011}
\item \cite{citeulike:11831265}
\item \cite{citeulike:7888211}
\item \cite{citeulike:478633}
\item \cite{opac-b1134063}
\item \cite{DBLP:conf/cscw/HowisonH11}
\item \cite{DBLP:conf/cscw/2011}
\item \cite{Trapani:2011}
\item \cite{Allsopp:2012}
\item \cite{1749-4699-6-1-015010}
\item \cite{Turk:2013:SCH:2484762.2484782}
\end{itemize}

@@

\appendix
\section{List of attendees \note{lead: Shel Swenson}}

@@

\bibliographystyle{unsrt}

\bibliography{wssspe}
\end{document}

>>>>

\begin{abstract}
This document discusses the First Workshop on Sustainable
Software for Science: Practice and Experiences (WSSSPE1).  The content
is based on the workshop itself, as recorded in a set of
collaborative notes taken during the workshop, including discussion
about the keynote presentations, three themed panels that focus on
papers accepted by the workshop, other discussion, cross-cutting
issues, use cases, and conclusions.  It joins a nascent literature
seeking to understand what drives software work in science and how the
reward systems of science thereby shapes the type of software work
undertaken, including the extent to which developers are motivated to
build software for the long-term, for the use of others, and whether
to work collaboratively or separately. Unique perspectives were
captured about issues such as documentation of APIs, software
deployment, and development practices; software licenses; career
paths for scientific software developers; identifiers for software
through a mechanism such as a Digital Object Identifier or
publication of a ``software paper''; and evidence of software
contribution and impact, recorded in online systems such as source
code repositories like GitHub. For truly sustainable software there
should be no endpoint, as the software products continue to be used
and useful beyond the single institution, grant, and developer or
development team.

@@

%\note{papers and presentations and URLs should be references - see
%  figshare-web in .bib file for URL style example}

@@

%\pagebreak
%
%\section*{Executive Summary}
%
%\todo{1 page summary goes here}
%
%\pagebreak

@@

Because progress in scientific research is dependent on the quality of and
accessibility to software at all levels, it is now critical to address many
challenges related to the development, deployment, and maintenance of reusable
software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists,
researchers, and students are able to learn and adopt software-related skills
and methodologies. Established researchers are already acquiring some of these
skills, and in particular a specialized class of software developers is
emerging in academic environments who are an integral and embedded part of
successful research teams. The WSSSPE workshop was intended to provide a forum
for discussion of the challenges, including both positions and experiences. The
short papers and debates have been archived to provide a basis for continued
discussion, and the workshop has fed into the collaborative writing of this
document. An estimate of 90 to 150 participants were present
throughout the day. Additional papers based on extended versions of workshop
submissions are expected. The level of interest in the workshop has led the
organizers, working with some of the submitters and attendees, to plan two
additional workshops (at the 2014 SciPy and SC14 conferences), to obtain
funds from the National Science Foundation and the Gordon and Betty Moore
Foundation to support these workshops, and to turn the
workshop website into a community website that can be used as a focus for
further discussion and progress. Additionally, a minisymposium that aims to
further explore some of the key issues raised in WSSSPE is co-organized by a
WSSSPE1 participant at the 2014 Society for Industrial and Applied Mathematics
(SIAM) Annual Meeting on ``Reliable
Computational Science.''

The goal of this paper is to summarize the contents of the workshop papers
and discussion, and identify issues that future WSSSPE events will address.
Before the WSSSPE1 workshop took place, the organizers self-published
a paper~\cite{WSSSPE1-pre-report} to document the process of
organizing and advertising the workshop, collecting and reviewing the
papers, and putting together the agenda. Section~\ref{sec:process} is
based on that publication. The remainder of this paper is based on the
workshop itself, as documented in a set of collaborative notes taken
during the workshop~\cite{WSSSPE1-google-notes}, including discussion
about the keynote presentations (\S\ref{sec:keynotes}), the three
panels (\S\ref{sec:devel}-\ref{sec:community}), 
cross-cutting issues (\S\ref{sec:cross-cutting}),
use cases (\S\ref{sec:use-cases}), and conclusions
(\S\ref{sec:conclusions}).

@@

The organizers decided to list 54 of the papers as significant
contributions to the workshop, a very high acceptance rate but one
that is reasonable, given the goal of broad participation and the fact
that the reports were already self-published. The papers were
grouped into three main categories, namely \emph{Developing and
Supporting Software}, \emph{Policy}, and \emph{Communities}. Each
subject area was associated with a panel and discussion at the
workshop.

@@

@@

Communities are extremely important in software projects, and both their
building and continued engagement need attention during the project life cycle.
Several of the submitted papers discussed how communities have been built around
projects, and what is needed to enable a project to grow. Among these are public
source code hosting, mailing lists, documentation, wikis, bug trackers, software
downloads, continuous integration, software quality dashboards, and of course,
a general web presence to tie all of these things together. Questions were posed
around users not answering the questions of other users. Several participants
offered counterexamples, such as mailing lists where developers do not participate
as much due to users being more active, or whether the ``core
team'' can end up setting unrealistic expectations by doing too much. Team Geek~\cite{opac-b1134063}
and Turk's paper on scaling code in the human dimension~\cite{Turk:2013:SCH:2484762.2484782}  discuss how development
lists that are welcoming to people actually have many more people contributing.

@@

@@

\todo{some text is needed here before the first subsection}

\subsection{Modeling Sustainability}

@@

%\todo{seems like definitions will be mentioned elsewhere in the paper?}
Each of the framework papers included a definition of sustainability, with many overlaps.  In essence the papers related the question to change over time, an idea well captured in the definition used by the UK's Scientific Software Sustainability Institute: ``software you use today will be available---and continue to be improved and supported---in the future,'' as quoted in~\cite{Venters_WSSSPE}. This idea was also expressed as software that continues to serve its users~\cite{Pierce_WSSSPE}.
See \S\ref{sec:cross-cutting} for more on the discussion of definitions of sustainability in the workshop.

@@

\subsection{Credit, Citation, Impact}

@@

\subsection{Implementing Policy}

@@

%There were a number of papers categorized under the Communities banner and so
%two presenters each summarized half of the submissions in this area.
\todo{an intro to the whole section is needed}

\subsection{Communities}

%This collection of papers was summarized by Karen Cranston.

@@

%\subsection{Communities Part 2}

%Nancy Wilkins-Diehr summarized 6 papers in this part.

@@

\subsubsection{What are communities?}

@@

\subsubsection{Challenges of community}
include metrics for community success (e.g., more external
contributors than internal).

\subsubsection{Admirable scientific software communities}
\begin{itemize}
\item example communities
\item their (quick) origin stories.
\end{itemize}

\subsubsection{Resources for learning about software communities}
\begin{itemize}
\item academic fields studying communities (and software communities).
\item courses on online communities
\item need for software carpentry module on organizing communities?
\item \cite{howison_scientific_2011}
\item \cite{citeulike:11831265}
\item \cite{citeulike:7888211}
\item \cite{citeulike:478633}
\item \cite{opac-b1134063}
\item \cite{DBLP:conf/cscw/HowisonH11}
\item \cite{DBLP:conf/cscw/2011}
\item \cite{Trapani:2011}
\item \cite{Allsopp:2012}
\item \cite{1749-4699-6-1-015010}
\item \cite{Turk:2013:SCH:2484762.2484782}
\end{itemize}

\subsection{Industry \& Economic Models}

\todo{text about papers and discussion in this part of the workshop}

\subsection{Education \& Training}

\todo{text about papers and discussion in this part of the workshop}

@@

\appendix
\section{Papers Accepted and Discussed at WSSSPE1}

\section*{Developing and Supporting Software}

\section*{Policy}

\section*{Communities, Models, and Education}

\subsubsection*{Communities}

\section{Attendees \note{lead: Shel Swenson}}

@@

\bibliographystyle{vancouver}

\bibliography{wssspe_paper}
\end{document}
